---
title: "Stat 341 Assignment 1"
output: pdf_document
date: "2022-09-26"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Question 1: Basic R Calculations

```{r}
3^4 # q1 a)
```

```{r}
log(100, base = 7) # 1.b)
```

```{r}
# 1.c)
x <- seq(1, 100) 
sum(sapply(x, function(x) {1/(x^2)}))
```

```{r}
100 %% 7 # 1.d)
```

```{r}
# 1.e) 
dx_steps <- 0.001
x_val <- seq(0, pi/2, by = dx_steps)
sum(sapply(x_val, function(x){ sin(x) * dx_steps }))
```

```{r}
# 1.f) 
f <- function(x) {
  return (dexp(x, rate = 1/2))
}

dx_steps <- 0.001
x_val <- seq(0, 3, by = dx_steps)
sum(sapply(x_val, function(x){ dexp(x, rate = 1/2) * dx_steps }))
```

```{r}
# 1.g) 
f <- function(x) {
  return (x^2 + 3)
}

dx_steps <- 0.0001
x_val <- seq(-2, 2, by = dx_steps)
sum(sapply(x_val, function(x){ f(x) * dx_steps }))
```

## Question 2: Comparing Spread Attributes

### 2a)

```{=tex}
\begin{align*}
  SD(\mathcal{P}+b) &= \sqrt{\frac{\sum_{u \in \mathcal{P}} \left( (y_u + b) - (\overline{y} + b) \right)^2}{N}} \\
                    &= \sqrt{\frac{\sum_{u \in \mathcal{P}} \left( (y_u - \overline{y} + b - b) \right)^2}{N}} \\ 
                    &=  SD(\mathcal{P})
\end{align*}
```

Hence, Standard Deviation is location invariant. 

```{=tex}
\begin{align*}
  a(\mathcal{P} + b) = MAD(\mathcal{P} + b) &= \operatorname*{median}_{u \in \mathcal{P}} ~ \left| y_u + b - (\operatorname*{median}_{u \in \mathcal{P}} ~ y_u + b)\right|. \\
                 &= \operatorname*{median}_{u \in \mathcal{P}} ~ \left| y_u - \operatorname*{median}_{u \in \mathcal{P}} ~ y_u + b - b\right| \\ 
                 &=  MAD(\mathcal{P} + b)
\end{align*}
```

Hence, Median Absolute Deviation is location invariant. 


```{r}
# 2.d) 
SD <- function(y) {
  return (sqrt(sum((y - mean(y))^2) / length(y)))
}

MAD <- function(y) {
  return (median(y - median(y)))
}
```

```{r}
# 2.e) 
set.seed(341)
pop = rexp(1000)

y_val <- seq(-1, 4, by=0.1)

sc <- function(pop, y, attr){
  N <- length(pop) + 1
  return (sapply(y, function(y) { (N * (attr(c(pop, y)) - attr(pop))) })) 
}

delta_sd <- sc(pop, y_val, SD)
delta_mad <- sc(pop, y_val, MAD)

plot(y_val, delta_sd, type="l", lwd = 2, 
     main="SC for std deviation and absolute deviation", ylab="sensitivity", xlab="y", 
     xlim=c(-1,4), ylim=c(-1, 1), col="blue")
lines(y_val, delta_mad, type="l", lwd = 2, main="Sensitivity curve for the median absolute deviation", col="red")

legend(x = "bottomright",          # Position
       legend = c("std deviation", "median absolution deviation"),  # Legend texts
       col = c("blue", "red"), # Line colors
       cex = 0.8,
       lwd = 2)   
```

## Question 3: Write a rounded-barplot-making function

```{r}
# 3.a) 
rounded.barplot <- function(x, xlab){
  table_x <- table(x)
  categories <- names(table_x)
  categories_frequencies <- as.numeric(table_x)
  
  plot.new()
  plot(NULL, type="n", xlim=c(0, 10*length(categories_frequencies)), ylim=c(0, max(categories_frequencies)), axes=FALSE, ann=FALSE)
  
  axis(2, at=seq(from=0, to=max(categories_frequencies), by=10))
  mtext(xlab, side=1, line=2)
  mtext("Frequency", side=2, line=3)
  
  x_semi <- seq(-4.5, 4.5, by=0.01)
  y_semi <- sqrt(20.25-x_semi^2)
  
  for (i in c(1: length(categories_frequencies))){
    rect(10*(i-1), 0, 10*i-1, categories_frequencies[i], col = "gray", border = "black")
    mtext(categories[i], 1, at=10*i-5, cex=0.85)
    polygon(x_semi + 4.5 + 10*(i-1), y_semi + categories_frequencies[i], col = "gray")
  }
}
```

```{r}
# 3.b) 
set.seed(12345)
flavours = c("Mango","Papaya","Banana","Coconut","Guava","Guarana","Durian","Cashew")
candies = sample(flavours, size=300, prob=(1:8)/sum(1:8), replace=TRUE)

rounded.barplot(candies, xlab="Candy Flavour")
```

## Question 4: R Analysis Question

```{r}
# 4.a) 
setwd("C:/Users/2baja/OneDrive/Desktop/STAT 341/A1")
apartment_eval <- read.csv("Apartment_Building_Evaluation.csv")

score_90 <- apartment_eval[,"SCORE"] >= 90
sum(score_90)
```

```{r}
# 4.b) 
davenport <- which(apartment_eval[,"WARDNAME"] == "Davenport")
davenport_apartments <- apartment_eval[davenport,]
davenport_apartments_sorted_addresses <- davenport_apartments[order(-davenport_apartments$SCORE),"SITE_ADDRESS"]
davenport_apartments_sorted_addresses[c(1:5)]
```

```{r}
# 4.c) 
unique_wardnames <- unique(apartment_eval[,"WARDNAME"])
sapply(unique_wardnames, function(name) { mean(apartment_eval[which(apartment_eval$WARDNAME == name), "SCORE"]) })
```

```{r}
# 4.d) 
plot(apartment_eval$YEAR_BUILT, apartment_eval$SCORE, pch = 1, col=adjustcolor("black", alpha = 0.8), xlab="Year built", ylab="Apartment Score")

unique_years <- unique(apartment_eval[,"YEAR_BUILT"])
average_score_by_year <- sapply(unique_years, function(year_built) { mean(apartment_eval[which(apartment_eval$YEAR_BUILT == year_built), "SCORE"]) })

lines(unique_years, average_score_by_year, pch = 18, col="red", type="b")

legend(x = "bottomleft",          # Position
       legend = c("building score", "average building score per year"),  # Legend texts
       col = c("black", "red"),           # Line colors
       cex = 0.75,
       pch = c(1, 19)) 
```

```{r}
# 4.e) 
influence_values <- function(pop, attribute){
  N <- length(pop)
  attribute_total_pop <- attribute(pop)
  
  return (sapply(1:N, function(x) { abs(attribute_total_pop - attribute(pop[-x])) }))
}

mean_influence <- influence_values(apartment_eval$SCORE, mean)

plot(1:length(apartment_eval$SCORE), mean_influence, xlab = "Observation Number", ylab = "Influence")
```

## Including Plots

You can also embed plots, for example:

```{r pressure, echo=FALSE}
plot(pressure)
```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
